Solution of Differential Algebraic Equations with Manifold Constraints on the Latent Space of an Autoencoder

### 2019-05-14 paper/paper.tex
more writing; ignoring latex temps

\title{Solution of Differential Algebraic Equations with Manifold Constraints
on the Latent Space of an Autoencoder}
\author{Alejandro Francisco Queiruga}
\date{2018-9-20}

\begin{document}
\maketitle
\begin{abstract}
We train an autoencoder to determine the lower-dimensional
representation of the EOS surface using modern deep learning techniques
without using any phase labels. The decoder end of the model is then
plugged back into the balance laws, eliminating the constrain equation.
\end{abstract}

@@

Part of the new modern software engineering framework involves replacing
human-written code with computer-written code. We are applying this
mindset from the computer-vision and advertisement dominated Artificial
Intelligence space into scientific computing.

@@

The original redescribe our problem as a differential algebraic equation
with a constraint: the balances of mass and energy are constrained by
the EOS. Because the EOS was originally derived from data, a data-driven
approach lends itself in which the EOS is a set of points in
temperature-pressure-density-enthalpy space. (In the multicomponent
cases there are more dimensions.)

@@

We train in two phases. The first step is minibatched stochastic
gradient descent on the entire deep network. After a few epochs of this,
we do a step of linear regression on the last layer of the network,
denote by \(W^{dec}\), which directly controls the values. The linear
system that needs to be solved is easily generated by assembling the
Hessian of the total loss function,
\begin{align}
 \mathbf{G} &= \sum_{x\in X}\frac{\partial L}{\partial W^{dec}} \\
\mathbf{H} &= \sum_{x\in X}\frac{\partial^2 L}{\partial (W^{dec})^2}
\end{align}
and solving the linear system for an update to the last parameters.

@@

To form the hessian and gradient for this step, we sum over all of the
training data. (The bias \(b^{dec}\) is included in \(W^{dec}\).) (The
contractive penalty term drops out because it does not depend on
\(W^{dec}\).)

@@

\begin{equation}
\frac{\partial R}{\partial q} \Delta q^{k+1} = R(q^k)
\end{equation}

The function \(R(q)\) has complex logic encoded in it; if-then
statements appear if the model used activations such as rectifiers.

@@

The user specifies any two of the values,

Note that the encoder only necessarily works when all four values are
correct. Thus, it is necessary to actually solve the nonlinear problem
on the decoder:

\[D(q^0)_i = s^0_i \quad \text{for specified}\; i\]

This equation has two rows. It is solved with Newton's method until
convergence.

\[K_ij = \frac{\partial D_i}{\partial q_j}  \quad \text{for specified}\; i, \text{for}\;j=1,2\]

\[K\Delta q = s^0_i-D(q)_i\]

The problem is not gaurunteed to converge, however, and often does not.
E.g., if the initial guess was in the liquid phase and the target point
in the EOS space is in a solid, the iteration will fail. To address this
issue, we perform multiple starts with an initial condition in each
phase and equilibrium, discard points that diverge, and then pick the
\(q\) that is closest to the two specified components.

It is possible that none of the initial guesses for this iteration
converge. There are two possiblities: the target state point is
incorrectly specified or out-of-range (e.g., negative density), or the
network is incapable of representing the desired material properties
(i.e., poorly trained).

@@

>>>>

\title{Solution of Differential Algebraic Equations with Manifold Constraints
on the Latent Space of an Autoencoder}
\author{Alejandro Francisco Queiruga}
\date{\today}

\begin{document}
\maketitle
\begin{abstract}
We train an autoencoder to determine the lower-dimensional
representation of the EOS surface using modern deep learning techniques
without using any phase labels. The decoder end of the model is then
plugged back into the balance laws, eliminating the constraint equation.
\end{abstract}

@@

Part of the new modern software engineering framework involves replacing
human-written code with computer-written code. We are applying this
mindset from industry problems in the Artificial
Intelligence space into scientific computing.

@@

We redescribe our problem as a differential algebraic equation
with a constraint: the balances of mass and energy are constrained by
the EOS. Because the EOS was originally derived from data, a data-driven
approach lends itself in which the EOS is a set of points in
temperature-pressure-density-enthalpy space. (In the multicomponent
cases there are more dimensions.)

@@

The dataset is split into a standard $2/3-1/6-1/6$ division into
training-testing-validation groups. We note that we can always
manufacture more datapoints. We train by alternating phases. The first
step is minibatched stochastic descent on the entire deep network, assembling the losses on minibatches of the data, $X_{mini}$,
\begin{equation}
  L(X_{mini}) = \sum_{x\in X_{mini}} L(x)
  \end{equation}
to perform the out-of-the-box ADAM optimizer on all of the
parameters in the networks. (The box is
TensorFlow.) After a few epochs of this,
the training algorithm alternates to a few steps step of linear regression on the last layer of the network,
denoted by \(W^{dec}\), which directly controls the magnitudes of the values.
(The bias $b^{dec}$ is included in $W^{dec}$.) The linear
system that needs to be solved is easily generated by assembling the
gradient and Hessian of the total loss function,
\begin{align}
 \mathbf{G}\left(X_{train}\right) &= \sum_{x\in X_{train}}\frac{\partial L}{\partial W^{dec}} \\
\mathbf{H}\left(X_{train}\right) &= \sum_{x\in X_{train}}\frac{\partial^2 L}{\partial (W^{dec})^2}
\end{align}
and solving the linear system for an update to the last parameters.

@@

To form the hessian and gradient for this step, we sum over all of the
training data. (The contractive penalty term drops out because it does
not depend on \(W^{dec}\).)

@@

\begin{equation}
\frac{\partial R}{\partial q} \Delta q^{k+1} = R(q^k)
\end{equation}
This equation is deceptively simple: the function \(R(q)\) has complex logic encoded in it; if-then
statements appear if the model used activations such as rectifiers.
However, the complex material behavior which traditionally requires
the vast majority of programming was implemented automatically by the
models; implementing equations XX, XX, and XX require only a few lines
of human-written code.

@@

Since the original manifold is 2D, the user only needs to specify two
initial values. In a multiphase solver such as TOUGH, the state needs
to be specified with the two primary variables plus the state tag: 
\begin{verbatim}
9e5 20 Aqu
0.5 20 AqG
\end{verbatim}
Note that the first primary variable means pressure in the first line
for a pure-liquid state,
but saturation in the second line for the liquid-gas equilibrium.
This makes sense inuitively, as this is how humans are taught, but
still leads to bugs and occasionally requires looking up values in
textbooks or the manual. (Is it $S_{gas}$ or $S_{aqu}$ on the second
line?) A fully fledged solver needs to check the user specified
initial conditions for erroneous inputs.

For LatentSim, the input system is not yet as robust as a
fully-fledged user-facing simulator, but the inputs are specified in
the $T p \rho h$ space instead of the $q$ space.
The user (or test problem) specifies any two of the four values in a
dictionary format,
\begin{verbatim}
p=9e5 T=20
rho=500 T=20
\end{verbatim}
Then, the simulator must determine $q(t=0)$ from these two
values. This sets up a system of two equations.

Note that the encoder only necessarily works when all four values are
correct. Thus, it is necessary to actually solve the nonlinear problem
on the decoder:
\[D(q^0)_i = s^0_i \quad \text{for specified}\; i\]
This equation has two rows indexing into the four state variables;
e.g. only the rows associated with $p$ and $T$ are solved if those are
specified, which is usually the case. The above equation is solved
with Newton's method. The two corresponding rows are cut out of
the tangent matrix to $D$,
\[K_{ij} = \frac{\partial D_i}{\partial q_j}  \quad \text{for specified}\; i, \text{for}\;j=1,2\]
and the following system of equations is solved for an update to $q$,
\[K\Delta q = s^0_i-D(q)_i\]
Updating $q$ is repeated until convergence.

The initial condition problem is not gaurunteed to converge, however,
and often does not.
For example, if the initial guess was in the liquid phase and the target point
in the EOS space is in a solid, the iteration will fail.
To address this issue, we perform multiple starts with an initial condition in each
phase and equilibrium, discard points that diverge, and then pick the
\(q\) that is closest to the two specified components. The multiple
start points are specified at EOS development time to form a set of
values that represent middle points for each phase,
\begin{equation}
\left\{s_{guess}\right\} = \left\{ \left(\begin{array}{c}
\bar{T}_{gas}\\
\bar{p}_{gas}\\
\bar{\rho}_{gas}\\
\bar{h}_{gas}
\end{array}\right),\left(\begin{array}{c}
\bar{T}_{liq}\\
\bar{p}_{liq}\\
\bar{\rho}_{liq}\\
\bar{h}_{liq}
\end{array}\right),\left(\begin{array}{c}
\bar{T}_{ice}\\
\bar{p}_{ice}\\
\bar{\rho}_{ice}\\
\bar{h}_{ice}
\end{array}\right)... \right\}
  \end{equation}
The multistart method could be specified without any prior knowledge
with random points, but this is a sufficiently robust method for
now. Each of the possible guess points is used to fill in the two
unspecified values as guesses and then passed into the encoder.
E.g. if $p$ and $T$ were specified, each possible $\bar{\rho}$ and
$\bar{h}$ combination from the above sets are filled in to provide
multiple guesses for $q$,
\begin{equation}
s^0 = \left(\begin{array}{c}
T^*(0)\\
p^*(0)\\
\bar{\rho}_{guess}\\
\bar{h}_{guess}
\end{array}\right);\quad q^0_{guess}=E\left(s^0\right)
 \end{equation}
 These sets of $q^0_{guess}$ are then iterated upon using the Newton's
 method procedure. We further add
 under-relaxation to the iteration,
 \begin{equation}
q^0_{k+1} =q^0_{k} + \alpha \Delta q_{k}
\end{equation}
with the under relaxation parameter usually set to $\alpha \approx
0.1$.

 It is further possible that none of the initial guesses for this iteration
converge. There are two possiblities: the target state point is
incorrectly specified or out-of-range (e.g., negative density), or the
network is incapable of representing the desired material properties
(i.e., poorly trained).

@@

The simulation inputs are the name of the equation of state, .e.g,
``water\_lg'' or ``water\_slgc'', and the name of the architecture
hyperparameters, e.g. ``Classifying\_pT\_1.0\_1,3,6,12,Sigmoid''.
In a mature code, the name of the architecture loaded would be hidden
from the end-user and would be a setting which the doman scientist and developer freezes
after checking the training process.
